An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-29 latex_files/ch2/2_2_LFI_background.tex
added references for LFI background

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic
of these models is that we can sample (simulate) data points,
but that we cannot evaluate the likelihood of a specific set of observations
$\data$. Formally, a simulator-based model is described as a
parameterised family of probability density functions
$\{ p(\yb|\thetab) \}_{\thetab}$, whose closed-form is either
unknown or computationally intractable to evaluate. Whereas evaluating
$p(\yb|\thetab)$ is intractable, sampling is
feasible. Practically, a simulator can be understood as a black-box
data generating machine $M_r$\footnote{The subscript $r$ in $M_r$ indicates the
  \textit{random} simulator. In the next chapters we will introduce
  $M_d$ witch stands for the \textit{deterministic} simulator.} that
produces samples $\yb$ in a stochastic manner for any given a set of parameters $\thetab$.% i.e.\ $M_r(\thetab) \rightarrow \yb$.

Simulator-based models provide considerable modelling freedom; any physical process that can be
conceptualised as a computer program of finite steps can be modelled
as a simulator-based model without any compromise. The modelling
freedom allows for any amount of hidden (unobserved) internal variables
or rule-based decisions. Hence, implicit models are often used to
model physical phenomena in the natural sciences such as genetics, epidemiology, or neuroscience, to name a few examples.
The modelling freedom, however, makes parameter inference difficult.

The main difficulty in parameter inference with implicit models is that the likelihood function
$L(\thetab) = p(\data|\thetab)$ is intractable. The methods that we consider below resort to the basic definition of the likelihood as the (rescaled) probability for a parameter of the model to generate data $\yb$ that is similar to the observed data $\data$. More formally, the likelihood $L(\thetab)$ equals
\begin{equation} \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(M_r(\thetab) \in \region(\data))
\end{equation}
%
where $c_\epsilon$ is a proportionality factor that depends on
$\epsilon$ and $\region(\data)$ is an $\epsilon$ region around $\data$ that is defined via a distance function $d$, i.e.\ $\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon \}$.

The basic characterisation of the likelihood in \eqref{eq:likelihood} highlights two sources of intractability: The first is the computation of the probability $\Pr(M_r(\thetab) \in \region(\data))$, the second is the limit of $\epsilon \to 0$.
Whilst the probability can be approximated with samples, this becomes computationally infeasible if $\epsilon$ is too small. Hence, a large class of inference methods for implicit models work with $\epsilon >0$, which leads to the approximate likelihood function  $L_{d, \epsilon}(\thetab)$,
\begin{equation} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data)), \quad \text{where  } \epsilon > 0.
\end{equation}
Most often in likelihood-free inference, we are interested in the posterior distribution of the parameters $\thetab$ given some observed data. The approximate likelihood function in \eqref{eq:approx_likelihood} implies the approximate posterior  $p_{d,\epsilon}(\thetab|\data)$, 
\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto L_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

>>>>

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample, or simulate, data points, but that we cannot
evaluate the likelihood of a specific set of observations
$\data$. Formally, a simulator-based model is a parameterised family
of probability density functions $\{ p(\yb|\thetab) \}_{\thetab}$,
whose closed-form is either unknown or computationally intractable to
evaluate. Whereas evaluating $p(\yb|\thetab)$ is intractable, sampling
is feasible. Practically, a simulator can be understood as a black-box
data generating machine $M_r$\footnote{The subscript $r$ in $M_r$
indicates the \textit{random} simulator. In the next chapters we will
introduce $M_d$ which stands for the \textit{deterministic}
simulator.} that generates samples $\yb$ in a stochastic manner for
any given a set of parameters $\thetab$.% i.e.\
$M_r(\thetab) \rightarrow \yb$.

Simulator-based models provide considerable modelling freedom; any
physical process that can be conceptualised as a computer program of
finite steps can be modelled as a simulator-based model without any
compromise. The modelling freedom allows for any amount of hidden
(unobserved) internal variables or rule-based decisions. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as genetics, epidemiology, or neuroscience, to
name a few examples. Further background on simulator-based models and
example applications can be found in the articles
by \citet{Gutmann2016, Lintusaari2017, Sisson2018, Cranmer2020}.

The modelling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. The main
difficulty is that the likelihood function $L(\thetab) =
p(\data|\thetab)$ is intractable. To better see the sources of the
intractability, and to address them, we go back to the basic
characterisation of the likelihood as the (rescaled) probability for a
parameter of the model to generate data $\yb$ that is similar to the
observed data $\data$. More formally, the likelihood $L(\thetab)$
equals
\begin{equation} \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(M_r(\thetab) \in \region(\data))
\end{equation}
where $c_\epsilon$ is a proportionality factor that depends on
$\epsilon$ and $\region(\data)$ is an $\epsilon$ region around $\data$
that is defined via a distance function $d$, i.e.\ $\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}$.

The basic characterisation of the likelihood in \eqref{eq:likelihood}
highlights two sources of intractability: The first is the computation
of the probability $\Pr(M_r(\thetab) \in \region(\data))$, the second
is the limit of $\epsilon \to 0$.  Whilst the probability can be
approximated with samples, this becomes computationally infeasible if
$\epsilon$ is too small. Hence, a large class of inference methods
work with $\epsilon >0$, which leads to the approximate likelihood
function $L_{d, \epsilon}(\thetab)$,
\begin{equation} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data)), \quad \text{where  } \epsilon > 0.
\end{equation}
Most often in likelihood-free inference, we are interested in the posterior distribution of the parameters $\thetab$ given some observed data. The approximate likelihood function in \eqref{eq:approx_likelihood} implies the approximate posterior  $p_{d,\epsilon}(\thetab|\data)$, 
\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto L_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

The approximation in \eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractabilities of the likelihood
function in \eqref{eq:likelihood}. Other strategies include modelling
the (stochastic) relationship between $\thetab$ and $\yb$, and its
reverse, or framing likelihood-free inference as a ratio estimation
problem \citep[see e.g.\ ][]{Blum2010, Wood2010, Papamakarios2016,
Thomas2016, Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both OMC and robust OMC that we introduce next
are based on the approximation in \eqref{eq:approx_likelihood}.
